### 2015-03-23 Kokkos_PG_Views.tex
Some corrections.

Views may have up to (at least) 8 dimensions, and any number of these
may be run-time or compile-time.  The only limitation is that the
run-time dimensions (if any) must appear first, followed by all the
compile-time dimensions (if any).  For example, the following are
valid three-dimensional View types:
\begin{itemize}
\item \lstinline!View<int***>!  (3 run-time dimensions)
\item \lstinline!View<int**[8]>!  (2 run-time, 1 compile-time)
\item \lstinline!View<int*[3][8]>!  (1 run-time, 2 compile-time)
\item \lstinline!View<int[4][3][8]>!  (3 compile-time)
\end{itemize}
and the following are \emph{not} valid three-dimensional View types:
\begin{itemize}
\item \lstinline!View<int[4]**>!
\item \lstinline!View<int[4][3]*>!
\item \lstinline!View<int[4]*[8]>!
\item \lstinline!View<int*[3]*>!
\item \lstinline!View<int*[3][8]>!
\end{itemize}
This limitation comes from the implementation of View using C++
templates.  View's first template parameter must be a valid C++ type.

@@

C++ lets users construct data types that may ``look like'' numbers in
terms of syntax, but do arbitrarily complicated things inside.  Some
of those things may not be thread safe, like unprotected updates to
global state.  Others may perform badly in parallel, like fine-grained
dynamic memory allocation.  As a result, Kokkos restricts the set of
data types \texttt{T} which may be the entries of a View.

Users may always construct a View whose entries are 
\begin{itemize}
\item built-in data types (``plain old data''), like \texttt{int} or
  \texttt{double}, or
\item structs of built-in data types.
\end{itemize}
The latter specifically forbids nontrivial destructors, constructors
with dynamic memory allocation, or virtual methods of any kind.  In
particular, entries must have size determined at compile time by the
\texttt{sizeof} operator.\footnote{Kokkos has a customization facility
  for users to add support for types with size determined at run time,
  but this is for expert users only.  For examples, see the Stokhos
  software package in the Trilinos project.}

@@

You may access an entry of a View using parentheses enclosing a
comma-delimited list of integer indices.  This looks just like a
Fortran multidimensional array access.  For example:
\begin{lstlisting}
const size_t N = ...;
View<double*[3][4]> a ("some label", N);
// KOKKOS_LAMBDA macro includes capture-by-value specifier [=].
parallel_for (N, KOKKOS_LAMBDA (const ptrdiff_t i) {
    const size_t j = ...;
    const size_t k = ...;
    const double a_ijk = a(i,j,k);
    /* rest of the loop body */
  });
\end{lstlisting}
Note how in the above example, we only access the View's entries in a
parallel loop body.  In general, you may only access a View's entries
in an execution space which is allowed to access that View's memory
space.  For example, if the default memory space is \lstinline!Cuda!
(CUDA device memory), you may not access it in host code, unless CUDA
UVM is enabled.  Furthermore, access costs (e.g., latency and
bandwidth) may vary, depending on the View's ``native'' memory and
execution spaces, and the execution space from which you access it.
CUDA UVM may work, but it may also be slow, depending on your access
pattern and performance requirements.  Thus, best practice is to
access the View only in a Kokkos parallel for, reduce, or scan, using
the same execution space as the View.  This also ensures that access
to the View's entries respect first-touch allocation.  The first
(leftmost) dimension of the View is the \emph{parallel dimension},
over which it is most efficient to do parallel array access.

@@

Views are allocated by default, in the default execution space's
default memory space.  You may access the View's execution space via
its \lstinline!execution_space! typedef, and its memory space via its
\lstinline!memory_space! typedef.  You may also specify the execution space
explicitly as a template parameter.  For example, the following
allocates a View in CUDA device memory:
\begin{lstlisting}
View<int*, Cuda> a ("a", 100000);
\end{lstlisting}
and the following allocates a View in ``host'' memory, using OpenMP
for first-touch allocation:
\begin{lstlisting}
View<int*, OpenMP> a ("a", 100000);
\end{lstlisting}

@@

You may allocate a View without initializing.  For example:
\begin{lstlisting}
View<int*> x (ViewAllocateWithoutInitializing (label), 100000);
\end{lstlisting}

@@

We discourage access to a View's ``raw'' pointer.  This circumvents
reference counting.  That is, the memory may be deallocated once the
View's reference count goes to zero, so holding on to a raw pointer
may result in invalid memory access.  Furthermore, it may not even be
possible to access the View's memory from a given execution space.
For example, a View in the \lstinline!Cuda! space points to CUDA device
memory.  Nevertheless, sometimes you really need access to the
pointer.  For such cases, we provide the \lstinline!ptr_on_device()!
method.  For example:
\begin{lstlisting}
// Legacy function that takes a raw pointer.
extern void legacyFunction (double* x_raw, const size_t len);

// Your function that takes a View.
void myFunction (const View<double*>& x) {
  // DON'T DO THIS UNLESS YOU MUST
  double* a_raw = a.ptr_on_device ();
  const size_t N = x.dimension_0 ();
  legacyFunction (a_raw, N);
}
\end{lstlisting}

@@

Strides may differ from dimensions.  For example, Kokkos reserves the
right to pad each dimension for cache or vector alignment.  You may
access all the dimensions of a View using the \lstinline!dimension! method,
and all the strides using the \lstinline!stride! method.  Both take a raw
integer array, and only fill in as many entries as the number of
dimensions.  For example:
\begin{lstlisting}
const size_t N0 = ...;
const size_t N1 = ...;
const size_t N2 = ...;
View<int***> a ("a", N0, N1, N2);
size_t dims[3]
a.dimension (dims); // fill 'dims' with dimensions
size_t strides[3]
a.strides (dims); // fill 'strides' with strides
\end{lstlisting}
You may also refer to specific dimensions:
\begin{lstlisting}
const size_t n0 = a.dimension_0 ();
const size_t n2 = a.dimension_2 ();
\end{lstlisting}

@@

We prefer that users let Kokkos determine a View's layout, based on
its execution space.  However, sometimes you really need to specify
the layout.  For example, the BLAS and LAPACK libraries only accept
column-major arrays.  If you want to give a View to the BLAS or
LAPACK, that View must be \lstinline!LayoutLeft!.  You may specify the
layout as a template parameter of View.  For example:
\begin{lstlisting}
const size_t N0 = ...;
const size_t N1 = ...;
View<double**, LayoutLeft> A ("A", N0, N1);

// Get 'LDA' for BLAS / LAPACK
int strides[2]; // any integer type works in stride()
A.stride (strides);
const int LDA = strides[1];
\end{lstlisting}
You may ask a View for its layout via its \lstinline!array_layout! typedef.
This can be helpful for C++ template metaprogramming.  For example
(C++11 only):
\begin{lstlisting}
template<class ViewType>
void callBlas (const ViewType& A) {
  typedef typename ViewType::array_layout array_layout;
  if (std::is_same<array_layout, LayoutLeft>::value) {
    callSomeBlasFunction (A.ptr_on_device (), ...);
  } else {
    throw std::invalid_argument ("A is not LayoutLeft");
  }
}
\end{lstlisting}

@@

\subsection{Random read Views}

Const Views are useful for declaring intent and helping you debug at
compile time.  They also open the door to other, optional
optimizations.  In particular, Kokkos lets you declare that you plan
to access a read-only array irregularly (less spatial locality), using
the optional \lstinline!RandomAccess! template parameter.  You may only use
this with a const View.  For example:
\begin{lstlisting}
const size_t N0 = ...;
View<int*> a_nonconst ("a", N0); // allocate nonconst View
// Assign to const, RandomAccess View
View<const int*, RandomAccess> a_ra = a_nonconst;
\end{lstlisting}
For a View whose execution space is \lstinline!Cuda!, access to a
\lstinline!RandomAccess! View may use CUDA texture fetches.  Texture
fetches are not cache coherent with respect to writes, which is why
you must use read-only access.  The texture cache is optimized for
noncontiguous access, since it has a shorter cache line than the
regular cache.

>>>>

Views may have up to (at least) 8 dimensions, and any number of these
may be run-time or compile-time.  The only limitation is that the
run-time dimensions (if any) must appear first, followed by all the
compile-time dimensions (if any).  For example, the following are
valid three-dimensional View types:
\begin{itemize}
\item \lstinline!View<int***>!  (3 run-time dimensions)
\item \lstinline!View<int**[8]>!  (2 run-time, 1 compile-time)
\item \lstinline!View<int*[3][8]>!  (1 run-time, 2 compile-time)
\item \lstinline!View<int[4][3][8]>!  (3 compile-time)
\end{itemize}
and the following are \emph{not} valid three-dimensional View types:
\begin{itemize}
\item \lstinline!View<int[4]**>!
\item \lstinline!View<int[4][3]*>!
\item \lstinline!View<int[4]*[8]>!
\item \lstinline!View<int*[3]*>!
\end{itemize}
This limitation comes from the implementation of View using C++
templates.  View's first template parameter must be a valid C++ type.

@@

C++ lets users construct data types that may ``look like'' numbers in
terms of syntax, but do arbitrarily complicated things inside.  Some
of those things may not be thread safe, like unprotected updates to
global state.  Others may perform badly in parallel, like fine-grained
dynamic memory allocation.  Therefore it is strongly advised to use 
only simple data types inside Kokkos Views.

Users may always construct a View whose entries are 
\begin{itemize}
\item built-in data types (``plain old data''), like \texttt{int} or
  \texttt{double}, or
\item structs of built-in data types.
\end{itemize}

While it is in principal possible to have Kokkos Views of arbitrary 
objects there are a number of restrictions. The \lstinline|T| type 
must have a default constructor and destructor. For portability reasons
\lstinline|T| is not allowed to have virtual functions and one is not 
allowed to call functions with dynamic memory allocations inside of 
kernel code. Furthermore assignment operators as well as default constructor
and deconstructor must be marked with \lstinline|KOKKOS_[INLINE_]FUNCTION|. 
Keep in mind that the semantics of the resulting View are a combination of the 
Views 'view' semantics and the behaviour of the element type.

@@

You may access an entry of a View using parentheses enclosing a
comma-delimited list of integer indices.  This looks just like a
Fortran multidimensional array access.  For example:
\begin{lstlisting}
const size_t N = ...;
View<double*[3][4]> a ("some label", N);
// KOKKOS_LAMBDA macro includes capture-by-value specifier [=].
parallel_for (N, KOKKOS_LAMBDA (const ptrdiff_t i) {
    const size_t j = ...;
    const size_t k = ...;
    const double a_ijk = a(i,j,k);
    /* rest of the loop body */
  });
\end{lstlisting}
Note how in the above example, we only access the View's entries in a
parallel loop body.  In general, you may only access a View's entries
in an execution space which is allowed to access that View's memory
space.  For example, if the default execution space is \lstinline!Cuda!, 
a View for which no specific Memory Space was fiven may not be accessed 
in host code \footnote{An exemption is if you specified for CUDA compilation
that the default memory space is CudaUVMSpace, which can be accessed from 
the host.}
Furthermore, access costs (e.g., latency and
bandwidth) may vary, depending on the View's ``native'' memory and
execution spaces, and the execution space from which you access it.
CUDA UVM may work, but it may also be slow, depending on your access
pattern and performance requirements.  Thus, best practice is to
access the View only in a Kokkos parallel for, reduce, or scan, using
the same execution space as the View.  This also ensures that access
to the View's entries respect first-touch allocation.  The first
(leftmost) dimension of the View is the \emph{parallel dimension},
over which it is most efficient to do parallel array access if the 
default memory layout is used (e.g. if no specific memory layout is 
specified).

@@

Views are allocated by default, in the default execution space's
default memory space.  You may access the View's execution space via
its \lstinline!execution_space! typedef, and its memory space via its
\lstinline!memory_space! typedef.  You may also specify the memory space
explicitly as a template parameter.  For example, the following
allocates a View in CUDA device memory:
\begin{lstlisting}
View<int*, CudaSpace> a ("a", 100000);
\end{lstlisting}
and the following allocates a View in ``host'' memory, using the default 
host execution space for first-touch initialization:
\begin{lstlisting}
View<int*, HostSpace> a ("a", 100000);
\end{lstlisting}

Since there is no bijective association between execution spaces and memory 
spaces, Kokkos provides a way to explicitly provide both to a View as a 
\lstinline|Device|. 
\begin{lstlisting}
View<int*, Device<Cuda,CudaUVMSpace> > a ("a", 100000);
View<int*, Device<OpenMP,CudaUVMSpace> > b ("b", 100000);
\end{lstlisting}
In this cas \lstinline|a| and \lstinline|b| will live in the same memory space,
but \lstinline|a| will be initialized on the GPU while \lstinline|b| will be 
initialized on the host.

@@

You may allocate a View without initializing.  For example:
\begin{lstlisting}
View<int*> x (ViewAllocateWithoutInitializing (label), 100000);
\end{lstlisting}
This is useful in situations where your dominant use of the View
exhibits a complicated access pattern. In this case it is best
to run the most costly kernel directly after initialization to 
execute the first touch pattern and get optimal memory affinity.

@@

We discourage access to a View's ``raw'' pointer.  This circumvents
reference counting.  That is, the memory may be deallocated once the
View's reference count goes to zero, so holding on to a raw pointer
may result in invalid memory access.  Furthermore, it may not even be
possible to access the View's memory from a given execution space.
For example, a View in the \lstinline!Cuda! space points to CUDA device
memory. Also using raw pointers would normally defeat the usability 
of polymorpic layouts and automatic padding. 
Nevertheless, sometimes you really need access to the
pointer.  For such cases, we provide the \lstinline!ptr_on_device()!
method.  For example:
\begin{lstlisting}
// Legacy function that takes a raw pointer.
extern void legacyFunction (double* x_raw, const size_t len);

// Your function that takes a View.
void myFunction (const View<double*>& x) {
  // DON'T DO THIS UNLESS YOU MUST
  double* a_raw = a.ptr_on_device ();
  const size_t N = x.dimension_0 ();
  legacyFunction (a_raw, N);
}
\end{lstlisting}
A user is in most cases also allowed to obtain a pointer to a specific 
element via the usual \lstinline|&| operator. For example

\begin{lstlisting}
// Legacy function that takes a raw pointer.
void someLibraryFunction (double* x_raw);

KOKKOS_INLINE_FUNCTION
void foo(const View<double*>& x) {
  someLibraryFunction(&x[3]);
}
\end{lstlisting}

This is only valid if a Views reference type is an \lstinline|lvalue|.
That property can be queried statically at compile time from the view through
its \lstinline|reference_is_lvalue| member.

@@

Strides may differ from dimensions.  For example, Kokkos reserves the
right to pad each dimension for cache or vector alignment.  You may
access the dimensions of a View using the \lstinline!dimension! method,
which takes the index of the dimension.

Strides are accessed using the \lstinline!stride! method.  It takes a raw
integer array, and only fills in as many entries as the rank of the View.
For example:
\begin{lstlisting}
const size_t N0 = ...;
const size_t N1 = ...;
const size_t N2 = ...;
View<int***> a ("a", N0, N1, N2);

int dim1 = a.dimension (1); // returns dimension 1
size_t strides[3]
a.strides (dims); // fill 'strides' with strides
\end{lstlisting}
You may also refer to specific dimensions without a runtime parameter:
\begin{lstlisting}
const size_t n0 = a.dimension_0 ();
const size_t n2 = a.dimension_2 ();
\end{lstlisting}

@@

We prefer that users let Kokkos determine a View's layout, based on
its execution space.  However, sometimes you really need to specify
the layout.  For example, the BLAS and LAPACK libraries only accept
column-major arrays.  If you want to give a View to the BLAS or
LAPACK library, that View must be \lstinline!LayoutLeft!.  You may specify the
layout as a template parameter of View.  For example:
\begin{lstlisting}
const size_t N0 = ...;
const size_t N1 = ...;
View<double**, LayoutLeft> A ("A", N0, N1);

// Get 'LDA' for BLAS / LAPACK
int strides[2]; // any integer type works in stride()
A.stride (strides);
const int LDA = strides[1];
\end{lstlisting}
You may ask a View for its layout via its \lstinline!array_layout! typedef.
This can be helpful for C++ template metaprogramming.  For example:
\begin{lstlisting}
template<class ViewType>
void callBlas (const ViewType& A) {
  typedef typename ViewType::array_layout array_layout;
  if (std::is_same<array_layout, LayoutLeft>::value) {
    callSomeBlasFunction (A.ptr_on_device (), ...);
  } else {
    throw std::invalid_argument ("A is not LayoutLeft");
  }
}
\end{lstlisting}

@@

Const Views often allow the compiler to optimize more agressively by allowing it
to reason about possible write conflicts and data aliasing. For example in a vector
update \lstinline|a(i+1)+=b(i) with skewed indexing it is safe to vectorize if
\lstinline|b| is a View of const data.

\subsection{Memory Traits}

Another way to get optimized data accesses is to specify a memory trait. 
These traits are used to declare intended use of the particular view of 
an allocation. For example a particular kernel might use a view only for 
streaming writes. By declaring that intention Kokkos can insert the 
appropriate store intrinsics on each architecture if available.

Currently 
In particular, Kokkos lets you declare that you plan
to access a read-only array irregularly (less spatial locality), using
the optional \lstinline!RandomAccess! template parameter.  You may only use
this with a const View.  For example:
\begin{lstlisting}
const size_t N0 = ...;
View<int*> a_nonconst ("a", N0); // allocate nonconst View
// Assign to const, RandomAccess View
View<const int*, RandomAccess> a_ra = a_nonconst;
\end{lstlisting}
For a View whose execution space is \lstinline!Cuda!, access to a
\lstinline!RandomAccess! View may use CUDA texture fetches.  Texture
fetches are not cache coherent with respect to writes, which is why
you must use read-only access.  The texture cache is optimized for
noncontiguous access, since it has a shorter cache line than the
regular cache.
